### 2017-06-14 sec-model.tex
Update for reviewer c comments; add reviewer c response.

Given a MOOC log, we can define a set $\mathbf{A}$ of actions that a
student can take at any given time. For example, an action $a \in
\mathbf{A}$ might be ``viewing lecture'', ``taking quiz'', or ``viewing
forum''. For each student in the course $\ell \in \mathbf{L}$, we then
extract a list of action sequences $\mathbf{S}_\ell$ that he or she
produced as observed in the log, where each sequence $\mathbf{s} \in
\mathbf{S}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_M)$ with
each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
paper we chose to denote the end of a sequence as occurring when no further
actions occur within a 10 hour window of time (and thus our sequences
roughly correspond to one day's worth of activity). This decision reflects
our desire to uncover latent state transition behavior at the granularity
of day-to-day behavior. A different segmentation strategy could be used to
uncover hour-by-hour behavior, or week-by-week behavior, and this depends
entirely on the desired time resolution one wishes to be exhibited in the
transitions between the latent states. Different segmentation strategies
will result in different underlying training data, and thus different
meanings behind the patterns that the 2L-HMM will extract.

Each sequence $\mathbf{s} \in \mathbf{S}_\ell$ is associated with a latent
state $u \in \{1,\ldots,K\}$ (where $K$ is a fixed constant picked in
advance). The actions within the sequence $\mathbf{s} = (a_1, a_2, \ldots,
a_M)$ are then modeled as a first-order Markov process conditioned upon $u$
where each action is drawn from a distribution conditioned upon the
previous action (except for the first which is sampled from an initial
starting distribution). We can write the parameters for the first-order
Markov model associated with latent state $u$ as $\lambda^{(u)} =
(\pi^{(u)}, A^{(u)})$ where $\pi^{(u)}$ indicates the initial probability
vector of length $|\mathbf{A}|$ and $A^{(u)}$ is an $|\mathbf{A}| \times
|\mathbf{A}|$ matrix indicating the transition probabilities between each
pair of actions from $\mathbf{A}$.

Thus, the probability of a sequence $\mathbf{s}$ of length $M$ given its
latent state $u$ is
\begin{equation}
  P(\mathbf{s} \mid \lambda^{(u)}) = P(a_1 \mid \pi^{(u)}) \prod_{i=2}^M P(a_i \mid
  a_{i-1}, A^{(u)})
\end{equation}
where $P(a \mid \pi^{(u)}) = \pi^{(u)}_a$ is the probability of starting
with action $a$ and $P(a_i \mid a_{i-1}, A^{(u)}) = A^{(u)}_{a_{i-1}, a_i}$
is the transition probability of moving from action $a_{i-1}$ to $a_{i}$
given that the model is currently in latent state $u$.

We can compute the likelihood of a list of action sequences
$\mathbf{S}_\ell$ of length $N$ for a student $\ell$ by marginalizing over
all possible latent state sequences $(v_1, \ldots, v_N) \in U$ as
\begin{equation}
  P(\mathbf{S}_\ell \mid \Lambda)
  = \sum_{(v_1, \ldots, v_N) \in U} \left(P(v_1 \mid \Lambda)
  P(\mathbf{s}_1 \mid \lambda^{(v_1)}) \times \prod_{i=2}^N P(v_i \mid
  v_{i-1}, \Lambda) P(\mathbf{s}_i \mid \lambda^{(v_i)})\right)
\end{equation}
where $\Lambda$ is the set of all model parameters.  In our model, we let
$\Lambda = (\pi, A, \lambda^{(1)}, \ldots, \lambda^{(K)})$ where $\pi$ and
$A$ are the parameters of a first-order Markov model over the latent states
and each $\lambda^{(i)}$ consists of the parameters for the first-order
Markov model over action sequences for latent state $i$. Thus $\pi$
(without superscripts) is an initial probability vector of length $K$ and
$A$ (without superscripts) is a $K \times K$ transition probability matrix,
analogous to the case with the individual first-order Markov model
parameters $\lambda^{(i)}$ for each latent state.

This can be seen as a modification of the traditional hidden Markov
model~\cite{Rabiner:1990:RSR} where instead of discrete observations (one
for each latent state transition) we have observations that take the form
of entire sequences $\mathbf{s}_i = (a_1, a_2, \ldots a_M)$ whose
probabilities are computed using another (non-hidden) Markov model
conditioned upon the latent state $u_i$.

\subsection{Parameter Estimation}
\label{sec:param-estimation}
To learn the parameters of our model, we may attempt to use maximum
likelihood estimation. Unfortunately, a closed-form solution does not
exist, so we must appeal to the EM algorithm~\cite{Dempster:1977:JRSS}. In
particular, we propose a minor modification of the Baum-Welch
algorithm~\cite{Rabiner:1990:RSR} which is used to learn the parameters for
hidden Markov models in an unsupervised setting. In the following sections,
we will provide a brief description of the original Baum-Welch algorithm,
and then describe our modification.

\subsubsection{Baum-Welch for Traditional HMMs}
In the traditional HMM formulation, we have $\Lambda = (\pi, A, B)$ where
$\pi$ is the initial probability distribution over the latent states,
$A$ is a $N\times N$ matrix indicating the latent state transition
probabilities, and $B$ is a $N \times V$ matrix indicating the probability
of generating a discrete observation symbol from a given latent state. The
Baum-Welch algorithm (also called the forward-backward algorithm) defines
two sets of variables $\alpha_t(i)$ called the forward variables and
$\beta_t(i)$ called the backward variables.

$\alpha_t(i) = P(o_1, \ldots, o_t, q_t = i \mid \Lambda)$ is the probability of
generating the sequence of observations $(o_1, o_2, \ldots, o_t)$ up to time
$t$ and arriving in state $i$ at that time. They are typically defined
using the following recursion:
\begin{itemize}
  \item $\alpha_1(i) = \pi_i b_i(o_1)$, the probability of starting in
    state $i$ ($\pi_i$) times the probability of generating the first
    observation $o_1$ from state $i$.
  \item $\alpha_{t+1}(i) = b_i(o_{t+1})\sum_{j=1}^N \alpha_t(j) A_{ji}$,
    the probability of generating the observation $o_{t+1}$ from state $i$
    times the probability that we arrive in state $i$ from any other
    previous state after generating all of the other observations.
\end{itemize}

Analogously, $\beta_t(i) = P(o_{t+1}, \ldots, o_{T} \mid q_t = i, \Lambda)$
is the probability of generating the \emph{rest of the sequence} given that
we are in state $i$ at time $t$. They are also defined using a recursion:
\begin{itemize}
  \item $\beta_T(i) = 1$
  \item $\beta_t(i) = \sum_{j=1}^N \beta_{t+1}(j) A_{ij} b_j(o_{t+1})$, the
    probability of transitioning to any state $j$ and generating the
    observation $o_{t+1}$ times the probability of generating the rest of
    the sequence given that we transitioned to state $j$.
\end{itemize}

Given the $\alpha$s and the $\beta$s, we can compute $\gamma_t(i)$, the posterior
probability of being in a given state $i$ at time $t$, and $\xi_t(i,j)$,
the posterior probability of going through a transition from state $i$ to
state $j$ at time $t$ as
\begin{equation}
  \gamma_t(i) = \frac{\alpha_t(i)\beta_t(i)}{\sum_{j=1}^N
  \alpha_t(j)\beta_t(j)}
\end{equation}
and
\begin{equation}
  \xi_t(i,j) = \frac{\alpha_t(i) A_{ij} b_j(o_{t+1}) \beta_{t+1}(j)}
  {\sum_{j=1}^N \alpha_t(j)\beta_t(j)}
\end{equation}
respectively. This is the E-step of the EM algorithm.

Given $\gamma_t(i)$ and $\xi_t(i,j)$ for each sequence $\mathbf{o}_k$ in
some training data of length $M$, we can update our model parameters
$\Lambda$ as
\begin{align}
  \pi_i &= \frac{\sum_{k=1}^M \gamma^{(\mathbf{o}_k)}_1(i)}{M},\\
  A_{ij} &= \frac{\sum_{k=1}^M \sum_{t=1}^T \xi^{(\mathbf{o}_k)}_t(i,j)}
  {\sum_{k=1}^M \sum_{j=1}^N \sum_{t=1}^T \xi^{(\mathbf{o}_k)}_t(i,j)},
  \text{ and }\\
  b_i(v) &= \frac{\sum_{k=1}^M \sum_{t=1, o_{k,t} = v}^T
  \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_{k,t})}_t(i)}.
\end{align}
This is the M-step of the EM algorithm.

\subsubsection{Baum-Welch for HMMs with Sequence Observations}
The major deviation of our model from the traditional HMM is that our
observations are themselves sequences. Thus, in all of the equations
presented in the previous section, we treat an observation $o_t =
\mathbf{s}_t$ at time $t$ as consisting itself of a list of actions $(a_1,
\ldots, a_M)$. Thus, instead of defining $b_i(v)$ as the probability that a
discrete observation \emph{token} $v$ is generated from state $i$, we have
\begin{equation}
  b_i(\mathbf{s}_t) = P(\mathbf{s}_t \mid \lambda^{(i)})
  = P(a_1 \mid \pi^{(i)}) \prod_{k=2}^M P(a_k \mid
  a_{k-1}, A^{(i)})
\end{equation}
where $b_i(\mathbf{s})$ now defines the probability of an action sequence
using a first-order (non-hidden) Markov model.

Fortunately, the recursions for the $\alpha$ and $\beta$ variables remain
the same, as do the definitions of $\gamma_t(i)$ and $\xi_t(i,j)$ in the
E-step. The only change is in the updating equations in the M-step,
where we replace the update for $b_i(v)$ by a pair of updates: one for
$\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. We denote the $m$-th action in the
$t$-th sequence for the $k$-th student as $o_{k,t,m}$. The two updates can
be written as
\begin{align}
  \pi^{(i)}_{a}
  &= \frac{\sum_{k=1}^M \sum_{t=1,a_1 = a}^T \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_k)}_t(i)}, \text{ and }\\
  A^{(i)}_{ab}
  &= \frac{\sum_{k=1}^M \sum_{t=1}^T
  \sum_{m=2,a_{m-1}=a \land a_m = b}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \sum_{m=2}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
\end{align}
where $a_m = o_{t,k,m}$ (for notational conciseness). Our modified EM
algorithm for 2L-HMMs is provided as part of the \meta/
toolkit~\citemeta{}.

>>>>

Given a MOOC log, we can define a set $\mathbf{A}$ of actions that a
student can take at any given time. For example, an action $a \in
\mathbf{A}$ might be ``viewing lecture'', ``taking quiz'', or ``viewing
forum''. For each student in the course $\ell \in \mathbf{L}$, we then
extract a list of action sequences $\mathbf{O}_\ell$ that he or she
produced as observed in the log, where each sequence $\mathbf{o} \in
\mathbf{O}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_T)$ with
each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
paper we chose to denote the end of a sequence as occurring when no further
actions occur within a 10 hour window of time (and thus our sequences
roughly correspond to one day's worth of activity). This decision reflects
our desire to uncover latent state transition behavior at the granularity
of day-to-day behavior. A different segmentation strategy could be used to
uncover hour-by-hour behavior, or week-by-week behavior, and this depends
entirely on the desired time resolution one wishes to be exhibited in the
transitions between the latent states. Different segmentation strategies
will result in different underlying training data, and thus different
meanings behind the patterns that the 2L-HMM will extract.

Each sequence $\mathbf{o} \in \mathbf{O}_\ell$ is associated with a latent
state $u \in \{1,\ldots,K\}$ (where $K$ is a fixed constant picked in
advance). The actions within the sequence $\mathbf{o} = (a_1, a_2, \ldots,
a_T)$ are then modeled as a first-order Markov process conditioned upon $u$
where each action is drawn from a distribution conditioned upon the
previous action (except for the first which is sampled from an initial
starting distribution). We can write the parameters for the first-order
Markov model associated with latent state $u$ as $\lambda^{(u)} =
(\pi^{(u)}, A^{(u)})$ where $\pi^{(u)}$ indicates the initial probability
vector of length $|\mathbf{A}|$ and $A^{(u)}$ is an $|\mathbf{A}| \times
|\mathbf{A}|$ matrix indicating the transition probabilities between each
pair of actions from $\mathbf{A}$.

Thus, the probability of a sequence $\mathbf{o}$ of length $M$ given its
latent state $u$ is
\begin{equation}
  P(\mathbf{o} \mid \lambda^{(u)}) = P(a_1 \mid \pi^{(u)}) \prod_{i=2}^T P(a_i \mid
  a_{i-1}, A^{(u)})
\end{equation}
where $P(a \mid \pi^{(u)}) = \pi^{(u)}_a$ is the probability of starting
with action $a$ and $P(a_i \mid a_{i-1}, A^{(u)}) = A^{(u)}_{a_{i-1}, a_i}$
is the transition probability of moving from action $a_{i-1}$ to $a_{i}$
given that the model is currently in latent state $u$.

We can compute the likelihood of a list of action sequences
$\mathbf{O}_\ell$ of length $N$ for a student $\ell$ by marginalizing over
all possible latent state sequences $(v_1, \ldots, v_N) \in U$ as
\begin{equation}
  P(\mathbf{O}_\ell \mid \Lambda)
  = \sum_{(v_1, \ldots, v_N) \in U} \left(P(v_1 \mid \Lambda)
  P(\mathbf{o}_1 \mid \lambda^{(v_1)}) \times \prod_{i=2}^N P(v_i \mid
  v_{i-1}, \Lambda) P(\mathbf{o}_i \mid \lambda^{(v_i)})\right)
\end{equation}
where $\Lambda$ is the set of all model parameters.  In our model, we let
$\Lambda = (\pi, A, B)$ where $\pi$ and
$A$ are the parameters of a first-order Markov model over the latent states
and $B = (\lambda^{(1)}, \ldots, \lambda^{(K)})$ where each $\lambda^{(i)}$
consists of the parameters for the first-order Markov model over action
sequences for latent state $i$. Thus $\pi$ (without superscripts) is an
initial probability vector of length $K$ and $A$ (without superscripts) is
a $K \times K$ transition probability matrix, analogous to the case with
the individual first-order Markov model parameters $\lambda^{(i)}$ for each
latent state. We have in total $K + K^2 + K(|\mathbf{A}| + |\mathbf{A}|^2)$
parameters to be estimated from our sequence data.

This can be seen as a modification of the traditional hidden Markov
model with categorical outputs~\cite{Rabiner:1990:RSR} where instead of
discrete observations (one for each latent state transition) we have
observations that take the form of entire sequences $\mathbf{o}_i = (a_1,
a_2, \ldots a_T)$ whose probabilities are computed using another
(non-hidden) Markov model conditioned upon the latent state $u_i$.

\subsection{Parameter Estimation}
\label{sec:param-estimation}
To learn the parameters of our model, we may attempt to use maximum
likelihood estimation. Unfortunately, a closed-form solution does not
exist, so we must appeal to the EM algorithm~\cite{Dempster:1977:JRSS}. In
particular, we propose a minor modification of the Baum-Welch
algorithm~\cite{Rabiner:1990:RSR} which is used to learn the parameters for
hidden Markov models in an unsupervised setting. In the following sections,
we will provide a brief description of the original Baum-Welch algorithm
for unsupervised parameter estimation for categorical valued hidden Markov
models, and then describe our modification to allow for parameter
estimation for our 2L-HMM modification for sequence valued observations.

\subsubsection{Baum-Welch for Traditional HMMs}
In the traditional HMM formulation with categorical outputs, we have
$\Lambda = (\pi, A, B)$ where $\pi$ is the initial probability distribution
over the latent states (of length $K$), $A$ is a $K\times K$ matrix
indicating the latent state transition probabilities, and $B = (b_1,
\ldots, b_k)$ is a length $K$ vector where each entry $b_i$ is a
probability distribution over the possible discrete output symbols from
$\mathbf{A}$.

The goal of the Baum-Welch algorithm (also called the forward-backward
algorithm) is to learn the values for the parameters $\Lambda$ from a
collection of observed sequences of values from $\mathbf{A}$. Concretely,
our training data $D = (\mathbf{o}^{(1)}, \ldots, \mathbf{o}^{(M)})$ is a
collection of $M$ sequences $\mathbf{o}^{(k)}$, each of which consists of $T_k$
symbols (observations) from $\mathbf{A}$. The Baum-Welch algorithm defines
two sets of variables $\alpha^{(\mathbf{o})}_t(i)$ called the forward
variables and $\beta^{(\mathbf{o})}_t(i)$ called the backward
variables~\cite{Rabiner:1990:RSR} for each sequence $\mathbf{o} \in
\mathbf{D}$.

$\alpha^{(\mathbf{o})}_t(i) = P(o_1, \ldots, o_t, v_t = i \mid \Lambda)$ is
the probability of generating the sequence of observations $(o_1, o_2,
\ldots, o_t)$ up to time $t$ and arriving in state $i$ at that time. They
are typically defined using the following recursion:
\begin{itemize}
  \item $\alpha^{(\mathbf{o})}_1(i) = \pi_i b_i(o_1)$, the probability of starting in
    state $i$ ($\pi_i$) times the probability of generating the first
    observation $o_1$ from state $i$.
  \item $\alpha^{(\mathbf{o})}_{t+1}(i) = b_i(o_{t+1})\sum_{j=1}^K
    \alpha^{(\mathbf{o})}_t(j) A_{ji}$, the probability of generating the observation
    $o_{t+1}$ from state $i$ times the probability that we arrive in state
    $i$ from any other previous state after generating all of the other
    observations.
\end{itemize}

Analogously, $\beta^{(\mathbf{o})}_t(i) = P(o_{t+1}, \ldots, o_{T} \mid v_t
= i, \Lambda)$ is the probability of generating the \emph{rest of the
sequence} given that we are in state $i$ at time $t$. They are also defined
using a recursion:
\begin{itemize}
  \item $\beta^{(\mathbf{o})}_T(i) = 1$
  \item $\beta^{(\mathbf{o})}_t(i) = \sum_{j=1}^K
    \beta^{(\mathbf{o})}_{t+1}(j) A_{ij} b_j(o_{t+1})$, the probability of
    transitioning to any state $j$ and generating the observation $o_{t+1}$
    times the probability of generating the rest of the sequence given that
    we transitioned to state $j$.
\end{itemize}

Given the $\alpha$s and the $\beta$s, we can compute
$\gamma^{(\mathbf{o})}_t(i)$, the posterior probability of being in a given
state $i$ at time $t$, and $\xi^{(\mathbf{o})}_t(i,j)$, the posterior
probability of going through a transition from state $i$ to state $j$ at
time $t$ as
\begin{equation}
  \gamma^{(\mathbf{o})}_t(i) =
  \frac{\alpha^{(\mathbf{o})}_t(i)\beta^{(\mathbf{o})}_t(i)}{\sum_{j=1}^K
  \alpha^{(\mathbf{o})}_t(j)\beta^{(\mathbf{o})}_t(j)}
\end{equation}
and
\begin{equation}
  \xi^{(\mathbf{o})}_t(i,j) = \frac{\alpha^{(\mathbf{o})}_t(i) A_{ij}
  b_j(o_{t+1}) \beta^{(\mathbf{o})}_{t+1}(j)} {\sum_{j=1}^K
  \alpha^{(\mathbf{o})}_t(j)\beta^{(\mathbf{o})}_t(j)}
\end{equation}
respectively~\cite{Rabiner:1990:RSR}. Computing these variables for each
sequence $\mathbf{o} \in D$ is the E-step of the EM algorithm.

Given $\gamma^{(\mathbf{o})}_t(i)$ and $\xi^{(\mathbf{o})}_t(i,j)$ for each
sequence $\mathbf{o} \in \mathbf{D}$, we can update our model parameters $\Lambda$
as
\begin{align}
  \pi_i &= \frac{\sum_{\mathbf{o} \in \mathbf{D}}
  \gamma^{(\mathbf{o})}_1(i)}{|\mathbf{D}|},\\
  A_{ij} &= \frac{\sum_{\mathbf{o} \in \mathbf{D}} \sum_{t=1}^T
  \xi^{(\mathbf{o})}_t(i,j)} {\sum_{\mathbf{o} \in \mathbf{D}} \sum_{j=1}^N
  \sum_{t=1}^T \xi^{(\mathbf{o})}_t(i,j)},
  \text{ and }\\
  b_i(a) &= \frac{\sum_{\mathbf{o} \in \mathbf{D}} \sum_{t=1, o_t = a}^T
  \gamma^{(\mathbf{o})}_t(i)} {\sum_{\mathbf{o} \in \mathbf{D}}
  \sum_{t=1}^T \gamma^{(\mathbf{o})}_t(i)}
\end{align}
respectively~\cite{Rabiner:1990:RSR}. This is the M-step of the EM
algorithm.

\subsubsection{Baum-Welch for HMMs with Sequence Observations}
The major deviation of our model from the traditional categorical-valued
HMM is that our observations are themselves sequences of actions from
$\mathbf{A}$ rather than individual tokens. We again denote our parameters
as $\Lambda = (\pi, A, B)$ where $\pi$ is the initial probability
distribution over the latent states (of length $K$), $A$ is a $K \times K$
matrix indicating the latent state transition probabilities, but $B =
(\lambda^{(1)}, \ldots, \lambda^{(K)})$ is now a vector of length $K$ where
each element $\lambda^{(i)}$ consists of the parameters for a first-order
Markov model over the action space $\mathbf{A}$. Recall that each
$\lambda^{(i)} = (\pi^{(i)}, A^{(i)})$ where $\pi^{(i)}$ is an initial
action distribution over the $|\mathbf{A}|$ available actions, and
$A^{(i)}$ is the $|\mathbf{A}| \times |\mathbf{A}|$ transition matrix
between those actions.

The goal of the Baum-Welch algorithm is still to learn the paramters
$\Lambda$ for our model. What differs from the categorical-valued HMM case
is that our data now consists of a collection of \emph{lists} of sequences,
rather than just a collection of sequences. This means that our observation
values $\mathbf{o}_t$ are now \emph{themselves} sequences of values from
$\mathbf{A}$, rather than just being a single token from $\mathbf{A}$.
Formally, our training data $\mathbf{D} = (\mathbf{O}^{(1)}, \ldots,
\mathbf{O}^{(|L|)})$, where each $\mathbf{O}^{(\ell)} = (\mathbf{o}_1,
\ldots, \mathbf{o}_{T_\ell})$ is itself a list of sequences. Each sequence
$\mathbf{o}_k \in \mathbf{O}^{(\ell)}$ consists of a list of actions $(a_1,
\ldots, a_{T_k})$, each of which is a member of $\mathbf{A}$.

In order to run the forward-backward algorithm for an element $\mathbf{O}
\in \mathbf{D}$ to compute the $\alpha$ and $\beta$ recursions like before,
we must define $b_i(\mathbf{o}_t)$ in this setting where $\mathbf{o}_t =
(a_1, \ldots, a_{T_t})$ is now a sequence instead of a single token. We
define it as follows:
\begin{equation}
  b_i(\mathbf{o}_t) = P(\mathbf{o}_t \mid \lambda^{(i)})
  = P(a_1 \mid \pi^{(i)}) \prod_{k=2}^{T_t} P(a_k \mid a_{k-1}, A^{(i)})
  = \pi^{(i)}_{a_1} \prod_{k=2}^{T_t} A^{(i)}_{a_{k-1},a_{k}}.
\end{equation}
Fortunately, the recursions for $\alpha^{(\mathbf{O})}_t(i)$ and
$\beta^{(\mathbf{O})}_t(i)$ remain the same, as do the definitions of
$\gamma^{(\mathbf{O})}_t(i)$ and $\xi^{(\mathbf{O})}_t(i,j)$, in the
E-step. We can simply substitute the new definition for the output
distribution $b_i(\mathbf{o}_t)$ in those equations.

The substantial change is in the updating equations in the M-step,
where we replace the update for $b_i(a)$ (which used to be a categorical
distribution) by a pair of updates for the Markov chain for state $i$: one
for $\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. The two updates can be written as
\begin{align}
  \pi^{(i)}_{a}
  &= \frac{\sum_{\mathbf{O} \in \mathbf{D}} \sum_{\mathbf{o}_t \in
  \mathbf{O},\mathbf{o}_{t,1} = a} \gamma^{(\mathbf{O})}_t(i)}
  {\sum_{\mathbf{O} \in \mathbf{D}} \sum_{\mathbf{o}_t \in \mathbf{O}}
  \gamma^{(\mathbf{O})}_t(i)}, \text{ and }\\
  A^{(i)}_{ab}
  &= \frac{\sum_{\mathbf{O} \in \mathbf{D}} \sum_{\mathbf{o}_t \in \mathbf{O}}
  \sum_{m=2,\mathbf{o}_{t,m-1}=a \land \mathbf{o}_{t,m} = b}^{|\mathbf{o}_{t}|}
  \gamma^{(\mathbf{O})}_t(i)}
  {\sum_{\mathbf{O} \in \mathbf{D}} \sum_{\mathbf{o}_t \in \mathbf{O}}
  \sum_{m=2,\mathbf{o}_{t,m-1}=a}^{|\mathbf{o}_{t}|} \gamma^{(\mathbf{O})}_t(i)}.
\end{align}
These updates can be understood as follows. $\pi^{(i)}_a$ is the
probability that a sequence being generated by state $i$ begins with action
$a$. $\gamma^{(\mathbf{O})}_t(i)$ gives the probability of generating the
sequence $\mathbf{o}_t$ from state $i$, so we simply aggregate this for all
sequences $\mathbf{o}_t$ where the first action is $a$. We then normalize
this distribution to sum to 1 across all possible actions $a \in
\mathbf{A}$ to obtain our new estimate for $\pi^{(i)}_a$.  $A^{(i)}_{ab}$
is the probability that a sequence being generated by state $i$ currently
at action $a \in \mathbf{A}$ transitions to action $b \in \mathbf{A}$.
Thus, we compute the expected number of times we observe a transition from
$a$ to $b$ in a sequence generated by state $i$, and we normalize this
distribution to sum to 1 across all possible actions $b \in \mathbf{A}$.
Our modified EM algorithm for 2L-HMMs is provided as part of the \meta/
toolkit~\citemeta{}.

### 2017-06-14 sec-related-work.tex
Update for reviewer c comments; add reviewer c response.

>>>>

For example, in part-of-speech tagging~\cite{Jurafsky:2009:SLP} the output
distributions are categorical (distributions over words from a fixed
vocabulary) and the latent states represent the part-of-speech category for
a word. In speech recognition~\cite{Huang:1990:HMM} the output
distributions might be mixtures of Gaussians to predict real-valued vectors
extracted from short windows of a speech signal. In the domain of
econometrics, \citet{Hamilton:1990:JoE} explores HMMs in the context of
``regime-switching.'' In this framing, the goal is to understand how
econometric data changes by modeling discrete changes in ``regime'' as
having an impact on the resulting real-valued vector data observed. The
``regimes'' are represented with some sort of model that can produce
real-valued vector data, such as a multivariate Gaussian or an
auto-regressive model. The analogy with HMMs is that a ``regime'' is a
latent state, and the characterization of the regime itself is the output
distribution for that latent state. Our model can be seen as such a
``regime-switching'' model where the output of the ``regimes'' that
students are switching between are discrete-valued \emph{sequences} (as
opposed to real numbers, vectors of real numbers, or categorical symbols)
and the model used to represent a specific ``regime'' is an (observable)
Markov chain over the observed student actions. We view the switching
between ``regimes'' as the first ``layer'' of our model, and the
transitioning behavior \emph{within} a ``regime'' between the actions a
student takes as the second ``layer'' of our model.
